\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-07-31 docs/parco13/paper.tex
Swap sections 3 and 4
Minor changes to text.

\newcommand{\comment}[1]{\textcolor{green}{[DAC: #1]}\xspace}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\dt}{\mathrm{d}t}
\newcommand{\clawpack}{{\sc Clawpack}\xspace}
\newcommand{\forestclaw}{Forestclaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}
\newcommand{\manyclaw}{Manyclaw\xspace}

@@

With the advent of many-core chips, such as GPUs and the MIC architecture, comes the opportunity to sustain unprecedented rates of floating point
operations at comparably high integration density and low cost.  These
architectures, however, require careful structuring of the data layout and
memory access patterns to exhaust their multithreading and vectorization
capabilities.

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  The presented work does not
construct a traditional high-order element but defines each element to be a
dense computational patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw, Berger:2011du}.
% \comment{Review 1 points out that the statement in red may not be what we are
% intending to say here.  Mainly the distinction between claiming better
% convergence then a high-order method (which I think is not what we are saying)
% and just stating that our elements are patches themselves as opposed to a
% single DOF.}
% CB: reworded the criticized statement above.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% \clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC: Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{\clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
toplevel unit & forest of octrees & patch of $m^d$ FV cells \\
atomic unit & octree leaf & one DOF in each cell \\
parallelization & MPI & threads (\manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
% CB: dropped dependencies.
% dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
Each leaf as the atomic unit of \pforest houses a toplevel unit of \clawpack.
The term cell is used to identify a single DOF within a \clawpack patch.  The
proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified, and performant codes.
% if the integration is done well
\item We preserve the separation between the mesh on one hand and the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf. \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I have added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed on.
\end{enumerate}

@@

A main challenge is presented by the fact that the patch neighborhood
is only known to \pforest.  This patch connectivity information needs
to be propagated to the numerical code in \forestclaw that implements
the interaction with neighbor patches via the use of a layer of (typically two)
ghost cells surrounding each patch.  To this end, we have designed an
interface that provides read-only access to the sequence of blocks, the list of
patches for each, and a constant-time lookup of neighbor patches (and their
relative orientation which is nontrivial between blocks).
% Since a neighbor patch has $m \ge 4$ cells in each direction, the required
% ghost cells are available by construction.
Suitably informed by \pforest, \forestclaw stores only the patches local to
each MPI rank.

@@

@@

The MPI layer is addressed from within \pforest and not exposed to the
\forestclaw code.  The order of leaves is maintained in \pforest according to a
space filling curve.  Each MPI rank has a local view on its own partition,
augmented where necessary with information about one layer of ghost leaves
(see \Fig{cubed3}).

@@

\alert{Thought: Should we swap sections 3. and 4.?}

For hyperbolic problems, we integrate the solution on a single uniform patch,
containing $m^d$ cells, using the wave propagation algorithm described by R. J.
LeVeque \cite{LeVeque:1997eg} and implemented in \clawpack \cite{le:2002,
clawpack}.  We assume a single degree of freedom per cell and reconstruct a
piecewise constant solution to obtain left and right states at cell edges.  At
each edge, we solve Riemann problems to obtain left and right going waves
propagating at speeds determined from the solution to the Riemann problem.  For
scalar advection, the speed of each wave is simply the local advection speed at
the cell interface.  For non-linear problems and systems, an approximate
Riemann solver, such as a Roe solver \cite{roesolver}, is typically used.  Since
much of the physics of an application can be contained in the Riemann solver,
\forestclaw adopts \clawpack's interface to Riemann solvers effectively allowing
problems solved with \clawpack to be solvable in \forestclaw.  In order to
achieve second order accuracy wave limiters are used to suppress spurious
oscillations at sharp gradients in the flow.

@@

\begin{figure}
\begin{center}
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_advance_strong-crop}
  \\[1ex]
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_exchange_strong-crop}
\end{center}
\caption{Strong scaling of MPI parallelism for advection on a uniform mesh.
  Top: run time of time integration.  Bottom: time spent in local and parallel
  neighbor exchanges.  The number of patches is $2^{2 \ell}$ at level $\ell$.
  The flat lines for the smaller runs are caused by having
  less patches than MPI processes, leaving some of them idle.  For each data point
  we use the maximum time required across all processes.}
\label{fig:uniscale}
\end{figure}%
We begin our analysis of Example 1 with uniform-mesh experiments run on TACC's
Stampede supercomputer at different levels of resolution.  First of all,
we use 1 to 16 MPI processes on one 16-core compute node, obtaining a strong
scaling efficiency for the time integration between 97\%--103\% depending on
the resolution (from levels 1 and 8).  Then, we examine processor counts from
16 to 256 using 16 cores per node.  The scaling behavior remains linear for
the time integration itself, and mostly linear for the local and parallel
neighbor exchanges which require about 10\% of the time integration run time;
see \Fig{uniscale}.

We have examined weak scalability as well by using four times as many processes
for each level inclease, from level 6 at 64 cores to level 10 at 4096.  The
exchange times stay consistently below 16\% of the time integration.  The time
in seconds per time integration step varies between .0578 and .0602,
which yields a parallel efficiency of 96\%.

\begin{table}
  \begin{center}
    \begin{tabular}{cccc|rr}
      \multicolumn{4}{c|}{strategy} & \multicolumn{2}{c}{wall clock time} \\
      mesh & remesh & partition & time step & $P = 16$ & $P = 256$ \\
      \hline
      uniform & none       & by count  & global   & 3961.  & 256.  \\
      \hline
          AMR & every step & by count  & global   &  252.  &  54.6 \\
          AMR & every 4    & by count  & global   &  178.  &  39.7 \\
      \hline
          AMR & every step & by count  & subcycle &   99.9 &  17.3 \\
          AMR & every 4    & by count  & subcycle &   87.2 &  14.0 \\
      \hline
          AMR & every step & by weight & subcycle &   95.7 &  18.2 \\
          AMR & every 4    & by weight & subcycle &   84.4 & 14.2 \\
      \hline
    \end{tabular}
  \end{center}
\caption{Comparison of different meshing strategies for the advection example
  with a fixed maximum level 8, run on 16 and 256 cores, respectively.  The
  wall clock time includes the whole run of the program beginning to end (file
  I/O disabled).  The adaptive runs converge to a minimum level of 3.}
\label{tab:swirlwall}
\end{table}
Next, we examine how we can reduce the overall wall clock time of a simulation
by switching from uniform refinement to AMR with the same maximum level,
effectively coarsening where high resolution is not beneficial.  For an example
with a sharp front as depicted in \Fig{results2d} we expect considerable
savings by AMR, which we confirm in \Tab{swirlwall}.  The wall clock times can
be reduced by up to a factor of 50 by combining AMR with subcycling.  The
adaptive runs are still very well load balanced overall, which we infer from
the linear scaling of the time integration alone (not shown) and from the fact
that the weighted partition does not influence the run times.  There is the
tradeoff however, at least in the current implementation, that the adaptive
runs suffer from the loss in scalability associated with the neighbor exchanges
that we discussed earlier.

>>>>

\newcommand{\comment}[1]{\textcolor{green}{[DAC: #1]}\xspace}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\dt}{\mathrm{d}t}
\newcommand{\clawpack}{{\sc Clawpack}\xspace}
\newcommand{\forestclaw}{ForestClaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}
\newcommand{\manyclaw}{Manyclaw\xspace}

\newcommand{\ignore}[1]{}

@@

With the advent of many-core chips such as GPUs and the MIC architecture
 comes the opportunity to sustain unprecedented rates of floating point
operations at comparably high integration density and low cost.  These
architectures, however, require careful structuring of the data layout and
memory access patterns to exhaust their multithreading and vectorization
capabilities.

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  The present work does not
construct a traditional high-order element but defines each element to be a
dense computational patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw, Berger:2011du}.
% \comment{Review 1 points out that the statement in red may not be what we are
% intending to say here.  Mainly the distinction between claiming better
% convergence then a high-order method (which I think is not what we are saying)
% and just stating that our elements are patches themselves as opposed to a
% single DOF.}
% CB: reworded the criticized statement above.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% \clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC: Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{\clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
toplevel unit & forest of octrees & patch of $m^d$ FV cells \\
atomic unit & octree leaf & one DOF in each cell \\
parallelization & MPI & threads (\manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
% CB: dropped dependencies.
% dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
Each leaf as the atomic unit of \pforest houses a toplevel unit of \clawpack.
The term cell is used to identify a single DOF within a \clawpack patch.  The
proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified, and performant codes.
% if the integration is done well
\item We preserve the separation between the mesh on one hand and the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf. \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I have added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
% \comment{It is probably more like M*m^d floating point values - with metric terms, velocity
% fields, multiple copies of the solution (needed for time stepping, re-taking times steps, etc).
% but the idea here is the same. What is a good way to say this? }

\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed on.
\end{enumerate}

@@

A main challenge is presented by the fact that the patch neighborhood
is only known to \pforest.  This patch connectivity information needs
to be propagated to the numerical code in \forestclaw that implements
the interaction with neighbor patches via the use of a layer of (typically two)
ghost cells surrounding each patch.  To this end, we have designed an
interface that provides \ignore{read-only} access to the sequence of blocks, the list of
patches per-block, and a constant-time lookup of neighbor patches (and their
relative and, in general, nontrivial orientation between blocks).
% \comment{DAC: What do you mean by read-only?  Neighbors get updated by the ghost cell
% exchange process...}
% Since a neighbor patch has $m \ge 4$ cells in each direction, the required
% ghost cells are available by construction.
Suitably informed by \pforest, \forestclaw stores only the patches local to
each MPI rank.

@@

\alert{I have swapped sections 3. and 4.  Please check/smooth the text.}

For hyperbolic problems, we integrate the solution on a single uniform patch,
containing $m^d$ cells, using the wave propagation algorithm described by R. J.
LeVeque \cite{LeVeque:1997eg} and implemented in \clawpack \cite{le:2002,
clawpack}.  We assume a single degree of freedom per cell and reconstruct a
piecewise constant solution to obtain left and right states at cell edges.  At
each edge, we solve Riemann problems to obtain left and right going waves
propagating at speeds determined from the solution to the Riemann problem.  For
scalar advection, the speed of each wave is simply the local advection speed at
the cell interface.  For non-linear problems and systems, an approximate
Riemann solver, such as a Roe solver \cite{roesolver}, is typically used.  Since
much of the physics of an application can be contained in the Riemann solver,
\forestclaw adopts \clawpack's interface to Riemann solvers effectively allowing
problems solved with \clawpack to be solvable in \forestclaw.  Wave limiters
are used to supress spurious oscillations resulting from truncation error in a typical
second order method.
\comment{DAC: Limiters formally reduce the order of accuracy.}
\ignore{In order to
achieve second order accuracy wave limiters are used to suppress spurious
oscillations at sharp gradients in the flow.}

@@

The MPI layer is addressed from within \pforest and not exposed to the
\forestclaw code.  The \ignore{order of leaves} leaf ordering
is maintained in \pforest according to a
space filling curve.  Each MPI rank has a local view on its own partition,
augmented where necessary with information about one layer of ghost leaves
(see \Fig{cubed3}).

@@

@@

\begin{figure}
\begin{center}
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_advance_strong-crop}
  \\[1ex]
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_exchange_strong-crop}
\end{center}
\caption{Strong scaling of MPI parallelism for advection on a
  uniformly refined mesh.  Top: run time of time integration.  Bottom:
  time spent in local and parallel neighbor exchanges.  The number of
  patches is $2^{2 \ell}$ at level $\ell$.  The flat lines for the
  smaller runs are caused by having fewer patches than MPI processes,
  leaving some of the processors idle.  For each data point we use the
  maximum time required across all processes. \comment{This last
    sentence is a bit unclear.  What do you mean exactly?}}
\label{fig:uniscale}
\end{figure}%
We begin our analysis of Example 1 with uniformly-refined mesh experiments run on TACC's
Stampede supercomputer at different levels of resolution.
We use 1 to 16 MPI processes on one 16-core compute node, obtaining a strong
scaling efficiency for the time integration between 97\%--103\% depending on
the resolution (from levels 1 and 8).  Then, we examine processor counts from
16 to 256 using 16 cores per node.  The scaling behavior remains linear for
the time integration itself, and mostly linear for the local and parallel
neighbor exchanges which require about 10\% of the time integration run time;
see \Fig{uniscale}.

We have examined weak scalability as well by using four times as many processes
for each level increase, from level 6 at 64 cores to level 10 at 4096.  The
exchange times stay consistently below 16\% of the time integration.  The time
in seconds per time integration step varies between .0578 and .0602,
which yields a parallel efficiency of 96\%.

\begin{table}
  \begin{center}
    \begin{tabular}{cccc|rr}
      \multicolumn{4}{c|}{strategy} & \multicolumn{2}{c}{wall clock time} \\
      mesh & remesh & partition & time step & $P = 16$ & $P = 256$ \\
      \hline
      uniform & none       & by count  & global   & 3961.  & 256.  \\
      \hline
          AMR & every step & by count  & global   &  252.  &  54.6 \\
          AMR & every 4    & by count  & global   &  178.  &  39.7 \\
      \hline
          AMR & every step & by count  & subcycle &   99.9 &  17.3 \\
          AMR & every 4    & by count  & subcycle &   87.2 &  14.0 \\
      \hline
          AMR & every step & by weight & subcycle &   95.7 &  18.2 \\
          AMR & every 4    & by weight & subcycle &   84.4 & 14.2 \\
      \hline
    \end{tabular}
  \end{center}
\caption{Comparison of different meshing strategies for the advection example
  with a fixed maximum level 8, run on 16 and 256 cores, respectively.  The
  wall clock time includes the whole run of the program beginning to end (file
  I/O disabled).  The adaptive runs converge to a minimum level of 3.}
\label{tab:swirlwall}
\end{table}
Next, we examine how we can reduce the overall wall clock time of a simulation
by switching from uniformly refined meshes to adaptive refinement with the same maximum level,
effectively coarsening where high resolution is not beneficial.  For an example
with a sharp front as depicted in \Fig{results2d} we expect considerable
savings by AMR, which we confirm in \Tab{swirlwall}.  The wall clock times can
be reduced by up to a factor of 50 by combining AMR with subcycling.  The
adaptive runs are still  load balanced overall, which we infer from
the linear scaling of the time integration alone (not shown) and from the fact
that the weighted partition does not influence the run times.  There is the
tradeoff however, at least in the current implementation, that the adaptive
runs suffer from the loss in scalability associated with the ghost patch exchanges
described earlier. Currently, ghost patch exchanges are made across all levels, even
if only two levels are involved in the exchange.
